Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 2.dev

### 2024-09-28 main.tex
Corrected speed sensor name, as per Felix's feedback.

\section{Methods}
%
\subsection{Bicycle}
%
We modified an electric bicycle (Grenoble C8 HMB, Royal Dutch Gazelle, Dieren,
The Netherlands) with a custom motor in the steering assembly capable of
applying up to 7~\si{\newton\meter} of torque between the head tube and steer
tube, see Figure~\ref{fig:balance-assist-bicycle}. A custom motor controller
converts the commanded torque to applied torque. We measure the speed of the
rear wheel with an encoder (ABS Speed Sensor, Magura GmbH, Bad Urach, Germany)
and measure the body fixed roll rate of the bicycle with a MEMs rate gyroscope
(VR IMU BN0086, Sparkfun, Niwot, USA). The balance assist control algorithm is
implemented on a microprocessor (Teensy, PJRC, USA) and data from all sensors
is logged with a CAN bus (CanEdge2, CSS Electronics, Aabyhoej, Denmark) at
least 100 Hz.

@@

\subsection{Perturbations}
%
We apply longitudinal forces to the ends of each handlebar from earth-fixed
posts using an adapted Bump'Em system~\citep{Tan2020} which we arrange with
four motors (EC-90 Flat, Maxon Group, Switzerland) working in cooperation,
Figure~\ref{fig:setup-diagram}. The four motors are programmed to apply a light
force at all times to keep the ropes taught and to track a commanded force
profile using a PID controller running on a microprocessor (Arduino Mega 2560,
Arduino, Italy). We measure the force applied by each motor at the handlebar
via four inline S-style load cells (Bosche GmbH, Damme, Germany). All were
rated for 250~\si{\newton} except the left rear load cell which was rated for
500~\si{\newton}. The commanded force profiles are designed to apply an
external pulsive torque to the front frame at magnitudes varying from
\SIrange{16}{160}{\newton\meter}. The four motors are arranged at the four
corners of a 1~\si{\meter} wide by 2~\si{\meter} long treadmill (Bonte
Technology B.V., Zwolle, The Netherlands) that can reach speeds up to
18~\si{\kilo\meter\per\hour}. The general design of this perturbation system is
described in detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and
the physical arrangement is shown in Figure~\ref{fig:participant-in-set-up}.
Our modifications relative to Van De Velde's design include simplifying the
Bump'Em motor controller with an inexpensive microcontroller and the use of a
simpler non-active safety harness.
%
% NOTE : I get a compile error without the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[height=65mm]{figures/setup-diagram.png}}
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[height=65mm]{figures/participant-in-set-up.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations and the various system components.}
  \label{fig:experimental-setup}
\end{figure}

@@

We record the order number of each perturbation \(j\), i.e. first, second,
third, \ldots, to measure how many perturbations a rider is subjected to before
the current perturbation. At the initiation of each perturbation we log the
instantaneous steer and roll angles to characterize the configuration of the
bicycle when perturbed. The gain factor setting on the balance assist controller
indicates if the assistance is off \(g=0\) or on at two different levels: low
\(g=8\) or high \(g=10\). A recovery from the perturbation is successful if the
person neither places their foot down onto the treadmill surface nor their
wheel of the bicycle exits the width of the treadmill belt. We record this as a
binary variable \(f\) for ``fall''. All measured variables are reported in
Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall Outcome'' and
    ``Perturbation Order Number'' are recorded per perturbation instance. The
    remaining measures are time varying during each trial.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain & \(g\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
    Fall Outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

>>>>

\section{Methods}
%
\subsection{Bicycle}
%
We modified a Grenoble C8 HMB electric bicycle (Royal Dutch Gazelle, Dieren,
The Netherlands) with a custom motor in the steering assembly capable of
applying up to 7~\si{\newton\meter} of torque between the head tube and steer
tube, see Figure~\ref{fig:balance-assist-bicycle}. A custom motor controller
converts the commanded torque to applied torque. We measure the speed of the
rear wheel with a DF30 Speed Sensor (Bosch eBike Systems, Reutlingen, Germany)
and measure the body fixed roll rate of the bicycle with a VR IMU BN0086 MEMs
rate gyroscope (Sparkfun, Niwot, USA). The balance assist control algorithm is
implemented on a Teensy microprocessor (PJRC, USA) and data from all sensors is
logged with a CanEdge2 CAN bus (CSS Electronics, Aabyhoej, Denmark) at least
100 Hz.

@@

\subsection{Perturbations}
%
We apply longitudinal forces to the ends of each handlebar from earth-fixed
posts using an adapted Bump'Em system~\citep{Tan2020} which we arrange with
four EC-90 Flat motors (Maxon Group, Switzerland) working in cooperation,
Figure~\ref{fig:setup-diagram}. The four motors are programmed to apply a light
force at all times to keep the ropes taught and to track a commanded force
profile using a PID controller running on an Arduino Mega 2560 microprocessor
(Arduino, Italy). We measure the force applied by each motor at the handlebar
via four inline S-style load cells (Bosche GmbH, Damme, Germany). All were
rated for 250~\si{\newton} except the left rear load cell which was rated for
500~\si{\newton}. The commanded force profiles are designed to apply an
external pulsive torque to the front frame at magnitudes varying from
\SIrange{16}{160}{\newton\meter}. The four motors are arranged at the four
corners of a 1~\si{\meter} wide by 2~\si{\meter} long treadmill (Bonte
Technology B.V., Zwolle, The Netherlands) that can reach speeds up to
18~\si{\kilo\meter\per\hour}. The general design of this perturbation system is
described in detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and
the physical arrangement is shown in Figure~\ref{fig:participant-in-set-up}.
Our modifications relative to Van De Velde's design include simplifying the
Bump'Em motor controller with an inexpensive microcontroller and the use of a
simpler non-active safety harness.
%
% NOTE : I get a compile error without the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[height=65mm]{figures/setup-diagram.png}}
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[height=65mm]{figures/participant-in-set-up.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations and the various system components.}
  \label{fig:experimental-setup}
\end{figure}

@@

We record the order number of each perturbation \(j\), i.e. first, second,
third, \ldots, to measure how many perturbations a rider is subjected to before
the current perturbation. At the initiation of each perturbation we log the
instantaneous steer and roll angles to characterize the configuration of the
bicycle when perturbed. The gain factor setting on the balance assist controller
indicates if the assistance is off \(g=0\) or on at two different levels: low
\(g=8\) or high \(g=10\). A recovery from the perturbation is successful if the
person neither places their foot down onto the treadmill surface nor their
wheel of the bicycle exits the width of the treadmill belt. We record this as a
binary variable \(f\) for ``fall''. All measured variables are reported in
Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall Outcome'' and
    ``Perturbation Order Number'' are recorded per perturbation instance. The
    remaining measures are time varying during each trial.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain Factor & \(g\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
    Fall Outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}
